---
title: "Predicting Classe: A Practical Machine Learning Project"
author: "Elmer C. Peramo"
date: "12/6/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background
Using devices such as *Jawbone Up, Nike FuelBand*, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how *much* of a particular activity they do, but they rarely quantify *how well they do it*. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).


## Initial Preparation and Data Analysis
Loading the necessary packages for this project:
```{r prep}
library(dplyr)
library(lubridate)
library(caret)
library(rpart)
library(rpart.plot)
library(rattle)
library(randomForest)
library(ggplot2)
```

## Data Loading
The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

Assuming that I have already downloaded the pml-training.csv and pml-testing.csv and have set the working directory to where I've saved these files:
```{r dload}
dtrain<- read.csv("pml-training.csv", na.strings = c("NA", "#DIV/0!", ""))
dtest<- read.csv("pml-testing.csv", na.strings = c("NA", "#DIV/0!", ""))
```

## Exploratory Data Analyses
### Dimensions
```{r dim}
dim(dtrain)
```
### Structure
```{r str}
str(dtrain)
```
### Samples
```{r head}
head(dtrain)
```
### Descriptive Stats
```{r summary}
summary(dtrain)
```
### Transformation
```{r date}
dtrain$cvtd_timestamp<- as.Date(dtrain$cvtd_timestamp, format = "%m/%d/%Y %H:%M")
dtrain$Day<-factor(weekdays(dtrain$cvtd_timestamp))
```

### Looking for insights
```{r insight}
table(dtrain$classe) 
prop.table(table(dtrain$classe)) 
table(dtrain$user_name) 
prop.table(table(dtrain$user_name)) 
```

### Cleaning
```{r clean1}
# Deleting unrelated fields
classe<- dtrain$classe
bad_col<- grepl("^X|timestamp|window", names(dtrain))
dtrain<- dtrain[, !bad_col]
dtrain<- dtrain[, sapply(dtrain, is.numeric)]
dtrain$classe<- classe
bad_col<- grepl("^X|timestamp|window", names(dtest))
dtest<- dtest[, !bad_col]
dtest<- dtest[, sapply(dtest, is.numeric)]

```

```{r clean2}
# Deleting fields with missing values
dtrain <- dtrain[, colSums(is.na(dtrain)) == 0]
dtest <- dtest[, colSums(is.na(dtest)) == 0] 
```

## Partition Data for Training and Testing
```{r partition}
set.seed(1125)
Trn <- createDataPartition(dtrain$classe, p=0.72, list=F)
train <- dtrain[Trn, ]
test <- dtrain[-Trn, ]
dim(train)
dim(test)
```

## Using ML Algorithms
Because of time and resource constraints, I'm just going to use two ML algorithms and choose between the two which to implment in the test set.
### Decision Trees
I used cross validation with k=6.
```{r tree}
train_c <- trainControl(method="cv", number=6)
c_tree <- train(classe~., data=train, method="rpart", trControl=train_c)
fancyRpartPlot(c_tree$finalModel)
tr_pred <- predict(c_tree,newdata=test)
cm_ctree <- confusionMatrix(test$classe,tr_pred)

# confusion matrix and accuracy
cm_ctree$table
cm_ctree$overall[1]
```

### Random Forest
```{r rf}
c_rf <- train(classe~., data=train, method="rf", ntree=8, importance=TRUE, prox=TRUE, trControl=train_c)
c_rf
fr_pred <- predict(c_rf,newdata=test)
cm_rf <-confusionMatrix(test$classe,fr_pred) 
cm_rf$table
cm_rf$overall[1]
```
### Getting a Single Tree
```{r rftree}
getTree(c_rf$finalModel, k=2)
```

### Class Centers
```{r rfcenters}
data <- classCenter(train[,c(1,2)], train$classe, c_rf$finalModel$prox)
data <- as.data.frame(data);data$classe <- rownames(data)
p <- qplot(roll_belt,pitch_belt,color=classe, data = train)
p + geom_point(aes(x=roll_belt,y=pitch_belt,col=classe),size=5,shape=4,data = data)
```



## Conclusion
Comparing the two algorithms, it clearly shows that Random Forest is the better model. So, I will use it to predict the values of classe in the untouched test data set.
```{r conclusion}

validation <- predict(c_rf,newdata=dtest)
validation
```